\documentclass[a4paper]{book}
\usepackage{alltt}
\usepackage{multind}
\usepackage{hyperref}
\usepackage{url}
\usepackage{xcolor,fancyvrb, mdframed}
\usepackage{amsmath}
\usepackage{varioref}
\usepackage{synttree, proof, centernot}
\usepackage[firstpage]{draftwatermark}
\usepackage{verbatim}
\usepackage[note]{marginote}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{dsfont}
\usepackage{float}
\usepackage{threeparttable}
\usepackage{fontspec}
\usepackage{mathtools}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{placeins}
\usepackage{tikz}
\usepackage{tikzsymbols}
\usetikzlibrary{arrows}
\usepackage{graphicx}
\usepackage{amsmath,mathdots,dsfont}
\usepackage{amssymb}
\usepackage{newunicodechar}
\usepackage{theorem}

\usepackage{texments}
%%% for movies by alectryon
\usepackage{movies/snippets/assets/alectryon}
\usepackage{movies/snippets/assets/tango_subtle}

\setcounter{tocdepth}{1}
\input{coqmacros}

%%% One hypothesis per line 
\makeatletter
\renewcommand{\alectryon@hyps@sep}{\alectryon@nl}
\makeatother


\SetWatermarkLightness{0.8}
\definecolor{lightorange}{rgb}{1,0.9,0.9}
\SetWatermarkScale{0.35}
\SetWatermarkText{Work in Progress}
\SetWatermarkColor{lightorange}

\author{Pierre Castéran\\ LaBRI, Univ. Bordeaux, CNRS UMR 5800, \thanks{With contributions by Yves Bertot, \'Evelyne Contejean,  Jérémy Damour, Florian Hatat, Pascal Manoury and Théo Zimmermann. The formalization of primitive recursive functions was originally authored by 
Russel O'Connor\cite{OConnor05}.}}
% \date{\today}
\date{August 2021}

\title{Hydra Battles and Cie}
\makeindex{coq}
\makeindex{maths}
\makeindex{hydras}
\makeindex{additions}
\makeindex{primrec}
\begin{document}
\maketitle




% \cleardoublepage




% % \mbox{~}
% % \clearpage
% % \newpage
% % \mbox{~}
% \vspace{10cm}
% \begin{quote} 
% { \Large {\it
% ``I start from one point and go as far as possible.'' } 

% John Coltrane.}
% \end{quote}


% \begin{verse}
% Hydra hydra hydram \\

% Hydrae hydrae hydra \\

% Hydrae hydrae hydras \\

% Hydrarum hydris hydris\\
% \end{verse}


\tableofcontents
 

%-------------------------------------------------------------------

\chapter{Introduction}

  

\vspace{16pt}

\section{Generalities}

Proof assistants are excellent tools for exploring the structure of mathematical proofs,
studying  which hypotheses are really needed, and which proof patterns are useful and/or
necessary. Since the development of a theory is represented as a bunch of computer files,
everyone is able to read the proofs with an arbitrary level of detail, or to play with the theory by writing alternate proofs or definitions.


Among all the theorems proved with the help of proof assistants like \coq{}~\cite{Coq,BC04}, \hol{}~\cite{HOL}, \isabelle{}~\cite{isabelle},  etc.,
several statements and proofs  share some interesting features:
\begin{itemize}
\item Their statements are easy to understand, even by non-mathematicians
\item Their proof requires some non-trivial mathematical tools
\item Their mechanization on computer presents some methodological interest.
\end{itemize}






This is obviously the case of the four-color theorem~\cite{fourcolors}  and the Kepler conjecture~\cite{flyspeck2015}. We do not mention impressive works like the proof of the odd-order theorem ~\cite{oddorderthm}, since understanding its statement requires a quite good mathematical culture.


In this document, we present two examples which seem to have the above properties.

\begin{itemize}
\item Hydra games (a.k.a. \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians:
L. Kirby and J. Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}. 
Although the mathematical contents of this 
paper are quite advanced, the rules of hydra battles are very easy to understand. There are now several sites on the Internet where you can find tutorials on hydra games, together with simulators you can play with. See, for instance, the blogpost and source code written by Andrej Bauer~\cite{bauer2008,BauerHydra}.



Hydra battles, as well as Goodstein sequences~\cite{goodstein_1944, KP82}
are a nice way to present complex termination problems.
The article by Kirby and Paris presents a proof of termination
based on ordinal numbers, as well as a proof that this termination is not
provable in Peano arithmetic. In the book dedicated to 
J.P. ~Jouannaud \cite{HommageJPJ}, N.~Dershowitz and G.~Moser  give a thorough survey on this topic~\cite{Dershowitz2007}.

Let us underline the analogy between hydra battles and interactive theorem proving. Hercules is the user (you!), and hydra's heads are the sub-goals: you may think that applying a tactic would solve a sub-goal, but it results often in the multiplication of such tasks.

\item In the second part, we are interested in computing $x^n$ with the least number of multiplications as possible. We use the notion of \emph{addition chains}~\cite{brauer1939,DBLP:journals/ipl/BerstelB87}, to generate efficient certified exponentiation functions.
\end{itemize}

\paragraph*{Warning:}

This document is \emph{not} an introductory text for \coq{}, and there are many aspects of this proof assistant that are not covered. 
 The reader should already have some basic experience with the \coq{} system. The Reference Manual and several tutorials are available on the \coq{} website~\cite{Coq}. The first chapters of textbooks like \emph{Interactive Theorem Proving and Program Development}~\cite{BC04}, \emph{Software Foundations}~\cite{SF} or  \emph{Certified Programming with Dependent Types} ~\cite{chlipalacpdt2011} will give you the right background.

%%%%% ICI

 
\subsection{Documenting theories with Alectryon}

Quotations of \coq{} source and answers are progressively replaced from copy-pasted \emph{verbatim} to automatically generated \emph{LaTeX} blocks, using Clément Pit-Claudel's \emph{Alectryon} tool~\cite{alectryonpaper, alectryongithub}.
Many thanks to Jérémy Damour and Théo Zimmermann who designed tools for maintaining consistency between the always evolving \coq{} modules and documentation written in \emph{LaTeX}.

Besides the guarantee of consistency between theories and documentation, we hope to give a corpus for experimenting new ways of documenting the implementation of non-trivial mathematics on a proof assistant.


At present, this document is a hybrid version: Chapter~\vref{chapter:primrec} is fully adapted to \emph{Alectryon}, and we just started to re-write Chapter~\vref{chapter:hydras}.




\subsection{Trust in our proofs}
\label{sect:trust-in-proofs}

Unlike mathematical literature, where definitions and proofs are spread out over many articles and books,
the whole proof is now inside your computer. It is composed from the \texttt{.v} files you downloaded and
parts of \coq's standard library. Thus, there is no ambiguity in our definitions and the premises of the theorems. Furthermore, you will be able to navigate through the development, using your favorite text editor or IDE, and some commands like \texttt{Search}, \texttt{Locate},  etc.



\subsection{Assumed redundancy}

It may often happen that several definitions of a given concept, or several proofs of a given theorem are possible. If all the versions present some interest, we will make them available, since each one may be of some methodological 
interest (by illustrating some tactic of proof pattern, for instance).
We use \coq's module system to make several proofs of a given theorem co-exist in our libraries (see also Sect~\vref{sect:alt-proofs}).
After some discussions of the pros and cons of each solution, we develop only one of them, leaving the others  as exercises or projects (i.e., big or difficult exercises).
In order to discuss which assumptions are really needed for proving a theorem, we will also present 
several aborted proofs.
Of course, do not hesitate to contribute nice proofs or alternative definitions!

It may also happen that some proof looks to be useless, because the proven theorem is a trivial consequence of another (also proven) result.
For instance, let us consider the three following statements:
\begin{enumerate}
\item There is no measure into $\mathbb{N}$ for proving the termination of all hydra battles (Sect~\vref{omega-case}).
\item There is no measure into the interval\footnote{We use the notation $[a,b)$ for denoting the set of ordinals greater or equal than $a$ and strictly less than $b$.}  $[0,\omega^2)$ for proving the termination of all hydra battles (Sect.~\vref{omega2-case}).
\item There is no measure into $[0,\mu)$ for proving the termination of all hydra battles, for any $\mu<\epsilon_0$ (Sect.\vref{sec:free-battles-case}).
\end{enumerate}

Obviously, the third theorem implies the second one, which implies the first one. So, theoretically, a library would contain only a proof of $(3)$ and remarks for $(2)$ and $(1)$. But we found it interesting to make all the three proofs available, allowing the reader to compare their common structure and notice their technical differences.
In particular, the proof of $(3)$ uses several non-trivial combinatorial properties of ordinal numbers up to $\epsilon_0$~\cite{KS81}, whilst $(1)$ and $(2)$ use simple properties of $\mathbb{N}$ and $\mathbb{N}^2$.


\subsection{About logic}

Most of the proofs we present are \emph{constructive}. Whenever possible, we provide the user with an associated function, which she or he can apply in \gallina{} or \ocaml{} in order to get a ``concrete'' feeling of the meaning of the considered theorem.
For instance, in Chapter~\vref{chap:ketonen}, the notion of \emph{limit ordinal} is
made more ``concrete'' thanks to a function \texttt{canon} which computes every item of a sequence which converges on a given limit ordinal $\alpha$. This simply typed function allows the user/reader to make her/his own experimentations.
For instance, one can very easily compute the $42$-nd item of a sequence which converges towards $\omega^{\omega^\omega}$.


 
Except in the \texttt{Schutte} library, dedicated to an axiomatic presentation of the set of countable ordinal numbers, all of our development is axiom-free, and respects the rules of intuitionistic logic. Note that we also use the \texttt{Equations} plug-in~\cite{sozeau:hal-01671777} in the definition of  several rapidly growing hierarchy of functions, in Chap.~\ref{chap:alpha-large}. This plug-in imports several known-as-harmless  axioms.

% \begin{Coqsrc}
% FunctionalExtensionality.functional_extensionality_dep : 
% forall (A : Type) (B : A -> Type) (f g : forall x : A, B x),
% (forall x : A, f x = g x) -> f = g
% \end{Coqsrc}

\index{coq}{Commands!Print Assumptions}

At any place of our development, you may use the  \texttt{Print Assumptions {\it ident}} command in order to verify on which axiom the theorem {\it ident} may depend. The following example is extracted from 
Library~\href{../theories/html/hydras.Epsilon0.F_alpha.html}{hydras.Epsilon0.F\_alpha}, where we use the \texttt{coq-equations} plug-in (see Sect.~\vref{sect:wainer}).

\begin{Coqsrc}
About F_zero_eqn.
\end{Coqsrc}

\begin{Coqanswer}
F_zero_eqn : forall i : nat, F_ Zero i = S i
\end{Coqanswer}

\begin{Coqsrc}
Print Assumptions F_zero_eqn. 
\end{Coqsrc}

\begin{Coqanswer}
Axioms:
FunctionalExtensionality.functional_extensionality_dep
  : forall (A : Type) (B : A -> Type) (f g : forall x : A, B x),
    (forall x : A, f x = g x) -> f = g
Eqdep.Eq_rect_eq.eq_rect_eq
  : forall (U : Type) (p : U) (Q : U -> Type) (x : Q p) (h : p = p),
    x = eq_rect p Q x p h
\end{Coqanswer}





\subsection{Typographical Conventions}

\subsubsection{Verbatim quotations}

Quotations from our  \coq{} source are displayed as follows:


  \begin{Coqsrc}
 Definition square (n:nat) := n * n.

 Lemma square_double : exists n:nat, n + n = square n.
 Proof.
    exists 2. 
  \end{Coqsrc}

Answers from \coq{} (including sub-goals, error messages, etc.) are displayed in slanted style
with a different background color.



 \begin{Coqanswer}
 1 subgoal, subgoal 1 (ID 5)
  
  ============================
   2 + 2 = square 2
   
 \end{Coqanswer}

 \begin{Coqsrc}
   reflexivity.
Qed.
 \end{Coqsrc}

In general, we do not include full proof scripts in this document. The only exceptions are very short proofs (\emph{e.g.}, proofs by computation, or by application of automatic tactics). Likewise, we may display only the important steps on a long interactive proof, for instance, in the following lemma (\vref{lemma:L-2_6-1}):

\begin{Coqsrc}
Lemma Lemma2_6_1 (alpha : T1) :  
  nf alpha -> forall beta,  beta t1< alpha  ->
  {n:nat | const_pathS n alpha beta}.
Proof.
  transfinite_induction alpha.
  (* ... *)
\end{Coqsrc}

The reader may consult the full proof scripts with Proof General or CoqIDE, for instance.

\subsubsection{Using Alectryon}

Alectryon's \emph{LaTeX} mode cannot be mistaken for our hand-made quotations, particularly if you read a color-print.
Here is an example from Chapter~\ref{chapter:primrec}.

\vspace{4pt}


\input{movies/snippets/Ack/AckFixpointIterate.tex}
\subsection{Active links}
The  links which appear in this pdf  document lead are of three possible kinds of destination:
\begin{itemize}
\item Local links to the document itself,
\item External links, mainly to \coq's website,
\item Local links to pages generated by \texttt{coqdoc}. According to the current makefile (through the commands \texttt{make html} and \texttt{make pdf}), 
  we assume that the page generated from a library \texttt{XXX/YYY.v} is stored as
the relative address \texttt{../theories/html/hydras.XXX.YYY.html} (from the location of the pdf)
Thus,  active links to our \coq{} modules may be incorrect if you did not get this \texttt{pdf} document by compiling the distribution available at
\url{https://github.com/coq-community/hydra-battles}.

\end{itemize}

\subsection{Alternative or bad definitions}
\label{sect:alt-proofs}
Finally, we decided to include definitions or lemma statements, as well as tactics,  that lead to
dead-ends or too complex developments, with the following coloring.
Bad definitions 
 are ''masked '' inside modules called \texttt{Bad}, \texttt{Bad1}, etc.


\begin{Coqbad}
Module Bad.

Definition double (n:nat)  := n + 2.
 
Lemma lt_double : forall n:nat, n < double  n.
Proof.
   unfold double; lia.
Qed.

End Bad.
\end{Coqbad}

Likewise, alternative, but still unexplored definitions will be presented in modules
\texttt{Alt}, \texttt{Alt1}, etc. Using these definitions is left as an implicit exercise.


\begin{Coqalt}
From hydras Require Import Iterates.
Module Alt.
  Definition double (n : nat) := iterate S n n.
End Alt.
\end{Coqalt}

\begin{Coqsrc}
Lemma alt_double_ok n : Nat.double n = Alt.double n.
Proof.
  unfold Alt.double, Nat.double; induction n; cbn.
  -  trivial.
  -  rewrite <- iterate_rw, iterate_S_eqn, <- IHn; lia.
Qed.
\end{Coqsrc}






\section{How to install the libraries}
\label{sec:orgheadline4}
\begin{itemize}
\item The present distribution has been checked with version 8.13.2 of the Coq proof assistant, with the plug-ins \texttt{coq-paramcoq}, \texttt{coq-equations}
and \texttt{coq-mathcomp-algebra}.

\item Please refer to \href{https://github.com/coq-community/hydra-battles#readme}{the README file of the project}
\end{itemize}

\section{Comments on exercises and projects}

Although we do not plan to include complete solutions to the exercises, 
we think it would be useful to include comments and hints, and questions/answers from the users. In contrast, ``projects'' are supposed, once completed, to be included in the repository.

Please consult the sub-directory \texttt{exercises/} of the
 project (in construction).

\section{Acknowledgements}
\label{sec:orgheadline5}
    Many thanks to Yves Bertot, \'Evelyne Contejean, Jéremy Damour,   Florian Hatat,  David Ilcinkas, 
Pascal Manoury,  Karl Palmskog, Cl\'ement Pit-Claudel, Sylvain Salvati, Alan Schmitt and Théo Zimmermann for their help on the elaboration of this library and  document, and to the
 members of the \emph{Formal Methods} team and the \emph{Coq working group} at laBRI for their helpful comments 
on  oral presentations of this work. 

Many thanks also to the Coq development team, Yves Bertot, and the members of the \emph{Coq Club} for interesting discussions about the \coq{} system and the Calculus of Inductive Constructions.

The author of the present document wishes to express his gratitude to the late Patrick Dehornoy, whose talk  was determinant for our desire to work on this topic.
I owe my interest in discrete mathematics and their relation to formal proofs and functional programming  to Srecko Brlek.  Equally, there is W. H. Burge's book ``\emph{Recursive Programming Techniques}'' ~\cite{burge} which was a great  source of inspiration.



\subsection{Contributions}

Yves Bertot made nice optimizations  to algorithms presented in Chapter~\ref{chapter-powers}.
\'Evelyne Contejean contributed libraries on the recursive path ordering (\emph{rpo}) for proving the well-foundedness of our representation of $\epsilon_0$ and $\Gamma_0$.
Florian Hatat proved many useful lemmas on countable sets, which we used in our adaptation of Schütte's formalization of countable ordinals. Pascal Manoury is integrating the ordinal $\omega^\omega$ into our hierarchy of ordinal notations.

The formalization of primitive recursive functions was originally a part of  Russel O'Connor's work on G\"odel's incompleteness theorems~\cite{OConnor05}. 

\label{sec:orgheadline2}

Any form of contribution  is welcome: correction of errors (typos and more serious mistakes), improvement of
Coq scripts, proposition of inclusion of new chapters, and generally any
comment or proposition that would help us. The text contains several \emph{projects} which, when completed, may improve the present work.
Please do not hesitate to share your contributions, for instance using pull requests and issues on GitHub. Thank you in advance!



% \subsubsection{Links to the Coq source}



\part{Hydras and ordinals}

\include{part-hydras}
\include{chapter-primrec}
\part{A few  certified algorithms}

\include{chapter-powers}

\part{Appendices}

\bibliographystyle{alpha}
\bibliography{thebib}





\chapter{Index and tables}

% \begin{todo}
%  Still very incomplete!
% \end{todo}
{\Large \textbf{In progress} This index is currently under reorganization for a few days. We apologize for its incompleteness! }

\printindex{coq}{Coq, plug-ins and standard library}
\printindex{maths}{Mathematical notions and algorithmics}
\printindex{hydras}{Library hydras: Ordinals and hydra battles}
\printindex{primrec}{Library hydras.Ackermann: Primitive recursive functions}
\printindex{additions}{Library additions: Addition chains}

\section{Main notations}






\begin{table}[H]
  \centering
  \begin{threeparttable}
    \caption{Ordinals and ordinal notations}
\begin{tabular}{|r | c|c|c|c|l|}
\hline
Name & Gallina&Math& Description& Page \\\hline
\texttt{lt : T1->T1->Prop}& lt alpha beta & $\alpha < \beta$& strict order on type \texttt{T1} \tnote{1} & \pageref{Predicates:lt-T1}\\
\texttt{LT: T1->T1->Prop}& alpha o< beta & $\alpha < \beta$& strict order on type \texttt{T1}   \tnote{2} & \pageref{Predicates:LT-T1}\\
\texttt{Lt : E0->E0->Prop} & alpha o< beta & $\alpha < \beta$& strict order on type \texttt{E0} \tnote{3} & \pageref{Predicates:Lt-E0} \\
\texttt{nf: T1->Prop} & \texttt{nf alpha} && alpha is in Cantor normal form & \pageref{Predicates:nf-T1}\\
 \texttt{on\_lt} & \texttt{alpha o< beta}&$\alpha<\beta$& ordinal inequality \tnote{4} & \pageref{sect:on-lt-notation}\\
 \texttt{on\_le} & \texttt{alpha o<= beta}&$\alpha\leq\beta$& ordinal inequality & \pageref{sect:on-lt-notation}\\
\texttt{plus} & \texttt{alpha + beta} & $\alpha + \beta$ & ordinal addition & \pageref{sect:infix-plus-T1}, \dots\\
\texttt{oplus} & \texttt{alpha o+ beta} & $\alpha \oplus \beta$ & Hessenberg sum & \pageref{sect:infix-oplus} \\

F & \texttt{F $n$} & $n$ & The $n$-th finite ordinal &  
\pageref{sect:notation-F}, \pageref{sect:notation-F-sch}\\ 
FS & \texttt{FS $n$} & $n+1$ & The $n+1$-th finite ordinal  \tnote{5} &  
\pageref{sect:notation-FS}\\ 
omega & \texttt{omega} & $\omega$ &   the first infinite ordinal   & \pageref{sect:notation-omega}, \pageref{sect:omega-T1}, \pageref{sect:omega-notation2}, \dots\\
phi0     & \texttt{phi0 alpha} & $\phi_0(\alpha),\; \omega^\alpha$&exponential of base $\omega$ & \pageref{sect:notation-phi0}\\

\hline
\end{tabular}
\begin{tablenotes}
  \item[1] This order is total, but not well-founded, because of not well formed terms.
\item[2] Restriction of \texttt{lt} to terms in normal form; this order is partial, but well-founded.
\item[3] This order is total \emph{and} well-founded.
\item [4]
Some notations may belong to several scopes. For instance, ``\texttt{o<}'' is
bound in \texttt{ON\_scope}, \texttt{E0\_scope}, \texttt{t1\_scope}, etc., and locally in several libraries.
  \item [5] Note that there exist also various coercions from \texttt{nat} to types of ordinal. Depending on the current scope and  \coq's syntactic analysis algorithm, \texttt{F} may be left implicit.
\end{tablenotes}
\end{threeparttable}
\end{table}



\vspace{4pt}


\begin{table}[H]

  \begin{threeparttable}
    \caption{hydra battles}
\begin{tabular}{|c|c|c|c|l|}
\hline
Name & Gallina&Math& Description& Page \\\hline
\texttt{round} & \texttt{h -1-> h'} & & one round of a battle & \pageref{sect:infix-round} \\
\texttt{rounds} & \texttt{h -+-> h'} & & one or more  rounds of a battle & \pageref{sect:infix-rounds} \\
\texttt{round\_star} & \texttt{h -*-> h'} & & any number of rounds of a battle & \pageref{sect:infix-rounds} \\
\hline
\end{tabular}

%\begin{tablenotes}
%\end{tablenotes}

  \end{threeparttable}
 
\end{table}

\begin{table}[H]
  \centering
  \begin{threeparttable}
    \caption{Addition chains}
\begin{tabular}{|c|c|c|c|l|}
\hline
Name & Gallina&Math& Description& Page \\\hline
\texttt{Mult} & \texttt{$z$ <--- $x$ times $y$} & & monadic notation & \pageref{monadic-mult} \\
\hline
\end{tabular}

%\begin{tablenotes}
%\end{tablenotes}

  \end{threeparttable}
 
\end{table}

\end{document} 
